---
title: "NSRAA compare survival from tow and control - coho"
output:
  html_notebook:
    theme: united
    toc: yes
  html_document:
    df_print: paged
    toc: yes
editor_options:
  chunk_output_type: inline
---

```{r setup, include=FALSE}
library(tidyverse)

knitr::opts_chunk$set(out.width = "100%", fig.width = 10, fig.height = 5)

# dirt <- "~/projects/nsraa_hatchery_release_stray"

```

# Data

Recovered tags included both rack and harvest.

```{r data, echo=FALSE}
tag_dat <- tibble(year = factor(rep(rep(c(2010, 2014, 2015, 2016), each = 2), 2)), # return year
                  group = rep(c("control", "tow", "control", "tow"), 4),
                  type = rep(c("recover", "total_tagged"), each = 8),
                  value = c(189, 228, 149, 163, 11, 165, 61, 65,
                            19117, 14657, 32734, 37418, 30608, 38784, 38746, 34205))

surv_dat <- tag_dat %>%
  pivot_wider(names_from = type, values_from = value) %>%
  mutate(dead = total_tagged - recover)

surv_dat %>% relocate(total_tagged, .after = group)

```


# $\chi^2$ test

```{r, echo=FALSE}
# automate chi-sq tests 
do_chisq <- function(dat_in, rtn_yr) {
  test_yr <- 
    dat_in %>%
    pivot_longer(c(recover,dead), names_to = "type") %>%
    select(-total_tagged) %>%
    filter(year == rtn_yr) %>%
    pivot_wider(names_from = group, values_from = value) %>%
    select(control, tow)
  chisq.test(test_yr, simulate.p.value = TRUE)
} # surv

rtn_yrs <- c(2010, 2014:2016)
chisq_out <- lapply(rtn_yrs, function(yr) do_chisq(surv_dat, yr))
names(chisq_out) = rtn_yrs

```

$\chi^2$ tables

2010
```{r, echo=FALSE}
print("Observed")
chisq_out$`2010`$observed

print("Expected")
chisq_out$`2010`$expected

```

2014
```{r, echo=FALSE}
print("Observed")
chisq_out$`2014`$observed

print("Expected")
chisq_out$`2014`$expected

```

2015
```{r, echo=FALSE}
print("Observed")
chisq_out$`2015`$observed

print("Expected")
chisq_out$`2015`$expected

```

2016
```{r, echo=FALSE}
print("Observed")
chisq_out$`2016`$observed

print("Expected")
chisq_out$`2016`$expected

```

*p*-values

```{r, echo=FALSE}
p <- sapply(chisq_out, function(out) out$p.value)
 
p %>%
  round(., 3) %>%
  ifelse(. < 0.001, " <0.001", .) %>%
  noquote

```

Combined *p*-value with Fisher's method

```{r, echo=FALSE}
metap::sumlog(p)

```


# Binomial regression model

I fitted a binomial model with proportions for dead/alive as the response variable and year, release method and the interaction term between year and release method as the explanatory variables. The intercept was removed in the model for easier interpretations.

```{r}
# model
surv <- glm(cbind(recover, dead) ~ year*group - 1 , family = binomial, data = surv_dat)

```

```{r, echo=FALSE}
# estimates and SE
est <- coef(surv)

# delta method
se <- msm::deltamethod(~ exp(x1+x5)/(1+exp(x1+x5)) - exp(x1)/(1+exp(x1)), est, vcov(surv))

diff <- plogis(est[1]+est[5]) - plogis(est[1])

ci <- diff + c(-1, 1)* qnorm(0.975)* se

```

The difference in survivals between the towed and control groups was estimated for each year. For example, in 2010, the difference in survivals between release methods was `r diff` (95% CI = (`r ci[1]`, `r ci[2]`)).

Below showed the model summary. The coefficients were in logit scale.

```{r}
summary(surv)

```

Table below shows the calculations for survival estimates.

| Return year   | Control   | Towed   |
| :------------ | :-------- | :------ |
| 2010          | $logit^{-1}(\beta_{yr2010})$ | $logit^{-1}(\beta_{yr2010} + \beta_{towed})$ |
| 2014          | $logit^{-1}(\beta_{yr2014})$ | $logit^{-1}(\beta_{yr2014} + \beta_{towed} + \beta_{yr2014\times towed})$ |
| 2015          | $logit^{-1}(\beta_{yr2015})$ | $logit^{-1}(\beta_{yr2015} + \beta_{towed} + \beta_{yr2015\times towed})$ |
| 2016          | $logit^{-1}(\beta_{yr2016})$ | $logit^{-1}(\beta_{yr2016} + \beta_{towed} + \beta_{yr2016\times towed})$ |

```{r, echo=FALSE}
# estimates and SE
est <- coef(surv)

# delta method
b2010c <- plogis(est[1]) %>% round(., 5)
se2010c <- msm::deltamethod(~ exp(x1)/(1+exp(x1)), est, vcov(surv)) %>% round(., 5)
b2010t <- plogis(est[1] + est[5]) %>% round(., 5)
se2010t <- msm::deltamethod(~ exp(x1+x5)/(1+exp(x1+x5)), est, vcov(surv)) %>% round(., 5)

b2014c <- plogis(est[2]) %>% round(., 5)
se2014c <- msm::deltamethod(~ exp(x2)/(1+exp(x2)), est, vcov(surv)) %>% round(., 5)
b2014t <- plogis(est[2] + est[5] + est[6]) %>% round(., 5)
se2014t <- msm::deltamethod(~ exp(x2+x5+x6)/(1+exp(x2+x5+x6)), est, vcov(surv)) %>% round(., 5)

b2015c <- plogis(est[3]) %>% round(., 5)
se2015c <- msm::deltamethod(~ exp(x3)/(1+exp(x3)), est, vcov(surv)) %>% round(., 5)
b2015t <- plogis(est[3] + est[5] + est[7]) %>% round(., 5)
se2015t <- msm::deltamethod(~ exp(x3+x5+x7)/(1+exp(x3+x5+x7)), est, vcov(surv)) %>% round(., 5)

b2016c <- plogis(est[4]) %>% round(., 5)
se2016c <- msm::deltamethod(~ exp(x4)/(1+exp(x4)), est, vcov(surv)) %>% round(., 5)
b2016t <- plogis(est[4] + est[5] + est[8]) %>% round(., 5)
se2016t <- msm::deltamethod(~ exp(x4+x5+x8)/(1+exp(x4+x5+x8)), est, vcov(surv)) %>% round(., 5)

```

The estimated survivals from the model.

| Return year   | Control (SE)  | Towed (SE)  |
| :------------ | :------------ | :---------- |
| 2010          | `r paste0(b2010c, " (", se2010c, ")")` | `r paste0(b2010t, " (", se2010t, ")")` |
| 2014          | `r paste0(b2014c, " (", se2014c, ")")` | `r paste0(b2014t, " (", se2014t, ")")` |
| 2015          | `r paste0(b2015c, " (", se2015c, ")")` | `r paste0(b2015t, " (", se2015t, ")")` |
| 2016          | `r paste0(b2016c, " (", se2016c, ")")` | `r paste0(b2016t, " (", se2016t, ")")` |























